A multi-scale story of the diffusion of a new technology: the Web

### 2025-09-28 paper/submission4/appendix.qmd
appendix

---
title: "A multi-scale story of the diffusion of a new technology: the Web"
author:
  - name: Emmanouil Tranos
    email: e.tranos@bristol.ac.uk
    affiliations: 
      - id: some-tech
        name: University of Bristol
        department: 
        address: 
        city: 
        state: State
        postal-code: UK
    corresponding: e.tranos@bristol.ac.uk
address:
  - attributes:
        corresponding: true
    note: This is the first author footnote.
abstract: |
    This paper investigates the spatial diffusion of a new technology that is the Web in the UK. It employs novel data and machine learning methods to model the influence of well-established diffusion mechanisms. Contrary to previous studies, it adopts multiple scales, high spatial granularity and a long study period that captures the early stages of the Web until its maturity (1996-2012). Findings reveal the importance of such spatial mechanisms (namely distance, urban hierarchy and the S-shaped pattern of the cumulative level of adoption) even at granular scales. They also highlight spatial heterogeneity and instances of leapfrogging. 
keywords: 
  - diffusion
  - technologies
  - Web
  - digital
date: last-modified
bibliography: bibliography
format:
  elsevier-pdf:
    keep-tex: true
    text: |
      \usepackage[demo]{graphicx} % "demo" option just for this example
      \usepackage{subcaption}
    journal:
      name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: authoryear
editor: 
  markdown: 
    wrap: 72
---

@@

# Introduction {#sec-introduction}

<!-- geographers lack of interest for diffusion + gap -->
Economic geographers were always interested in how new technologies and
innovations diffuse across space and time with the work of
@hagerstrand1968innovation, which demonstrated the importance of
communications as a mechanism for the diffusion of innovations, being
the most impactful. It was followed by various empirical studies -- see
for instance @ormrod1990 and @iso2005. However, the torch of exploring
and modelling such diffusion processes had been passed to other
disciplines such as economics, business studies and sociology well
before the 'cultural turn' of economic geography
[@perkins2005international; @ding2010modeling].
<!-- NEW spatial granularity --> As a consequence, relevant studies
did not investigate diffusion mechanisms at granular geographical scales to 
provide local policy insights, for instance, regarding infrastructural 
development and local technology policies. 
A potential explanation of the lack of geographical studies exploring the 
diffusion of new and, more specifically for this paper, digital technologies 
across both space and time can be attributed to the scarcity of relevant and 
granular enough data, a problem also highlighted by @iso2005 and
@kemeny2011international. As @zook2022mapping pointed, digital
activities are hardly ever captured in official data.

<!-- contribution -->
This paper offers such a contribution: an economic geography study
revealing the importance of spatial mechanisms for the diffusion of a
digital technology that is the Web in the UK at a high level
of spatial granularity 
<!-- NEW TO JUSTIFY DATA TIME PERIOD-->
from the very early days until its maturity (1996-2012).
<!-- NEW DATA CLARIFICATION -->
It specifically focuses on the diffusion of commercial websites.
It does so by employing a novel source of big data and machine learning
algorithms to model the *active* engagement with Web -- that is creating and 
maintaining instead of just browsing a website -- during that period
and tests how different spatial and temporal mechanisms shaped its
diffusion [@tranosuk]. This paper also exemplifies how the combination of data
sources and state-of-the-art research methods which escape the
traditional social science domain can offer new lenses to geographical
research regarding the understanding of technological diffusion.

<!-- motivation: stakeholders -->
The motivation for this paper lies in the fact that there are various
stakeholders who are interested in knowing how new digital technologies
diffuse over space and time and use this knowledge to make predictions
about the diffusion of related future technologies. As per
@leibowicz2016representing, historical studies agree that technologies
diffuse differently in terms of times, rates, and geographies and can be
driven by related policies [@victor1993]. @meade2021modelling highlight
that a variety of actors have a direct interest in gaining such
knowledge including network equipment suppliers, network operators,
regulatory and local authorities. These processes and their effects vary
a lot across scales: although the diffusion of a new technology might
not be optimal at a local level, it might be beneficial from a global
perspective as it could lead to faster diffusion to less advantaged
places [@leibowicz2016representing]. Despite the spatial heterogeneity
of such diffusion mechanisms and the policy relevance, there is a
scarcity of studies analysing the diffusion of new technologies
at a detailed geographical level because of the lack of detailed enough
data and adequate research methods.

<!-- motivation: regional development -->
This gap becomes even more prominent if we consider the importance of
technology and its adoption in economic geography and macroeconomic
growth theories. Simply put, differences in technological adoption and
sophistication may lead to uneven development as new technologies can
increase productivity and allow firms to introduce new products
[@solow1957technical; @aghion1990model; @kemeny2011international].
Linked to the subject of this paper, the Web and, more specifically, 
commercial websites can be considered as the underpinning
technology for the transformative process identified by
@capello2024nexus as the *digital service economy*. Importantly, the
focus here is not the Web in general, but instead commercial websites.
After all, diffusion together with invention and innovation are
considered the pillars of technological change [@das2022diffusion].
Findings from previous studies illustrated that early engagement with
digital technologies such as the Web led to longstanding positive
productivity gains for these regions that hosted early adopters
[@tranosuk]. Hence, obtaining a robust understanding of the spatial
diffusion mechanisms is crucial in mapping future development trajectories.

<!-- NEW motivation: intangible/digital -->
Another line of motivation has to do with the very nature of the Web as a 
digital technology and how it evolved over time. The digital geographies field 
has debunked premature claims about the Internet's placeless-ness by mapping 
various of its facets and technological layers [@tranos2013death], but the lack 
of relevant long time-series data prevented researchers from observing digital
phenomena overtime. Most of these studies focused on its hard infrastructure 
[e.g. @malecki2002economic; @moss2000internet], on Internet subscribers 
[e.g. @blank2018local], or on social media users [e.g. @crampton2013beyond].
Exceptions include early work from @zook2000web, @zook2001old and 
@moss1997tracking, which utilised domain name registration data and use it to
map and illustrate the spatial footprint of the web content. Pivotal as these 
contributions have been in understanding the economic geographies of the 
Internet, they were based on cross-sectional, often spatially detailed, mapping 
and on short temporal comparisons (e.g. 1993-1998) and did not observe the 
evolution and diffusion of the Web until its maturity. As we know from the 
relevant literature, different technologies diffuse at different rates and 
there is no theory that can explain these differences 
[@leibowicz2016representing]. Hence, mapping and modelling the diffusion of the
Web can offer useful insights.

<!-- R2 addition -->
The lack of previous granular studies about the diffusion of new technologies 
deprives us from a reliable prior regarding whether the diffusion mechanisms 
and drivers are sensitive or not to the spatial scale that the diffusion of a 
new technology is observed. As per other key concepts in economic geography
[@briant2010dots; @meliciani2015determinants; @andini2019social],
diffusion mechanisms can indeed suffer from Modifiable Areal Unit Problem 
(MAUP): although the adoption of a new technology is at its core an individual 
process, it is hardly ever observed at this, or at a close enough, scale. 
<!-- response to editor -->
Hence, this paper exposes the importance of spatially-explicit diffusion 
mechanisms at granular geographical scales without the risk of MAUP. The results
are linked to core debates in economic geography about agglomeration 
externalities, centripetal forces and the spatially clustered demand for new 
technologies. Although the paper focuses on the Web, the framework offered here 
and the findings which illustrate the micro-geographical patterns of the 
diffusion of a new technology are applicable to other new technologies.

<!-- early engagement with the web -->
This paper focuses on the diffusion of the Web as new technology during
the 1996-2012 period. This was an exciting period for digital
technologies as it corresponds with the commercialisation of the
Internet and, consequently, its almost universal adoption. The reader is
reminded that it was only in 1994 when Netscape Navigator was
introduced, a year before Microsoft's Internet Explorer.[^1] Also, only
9 per cent of UK's households had access to the Internet in 1998
[@ons2018], the web included mostly static webpages, there were no
social media and web browsing involved exclusively a desktop PCs as
there were no smartphones [@tranosuk]. Hence, it is fair to say that the
study period captures the very early stages of the diffusion of a new
technology that is the Web until its maturity. The former is a key point
in the lifecycle of a new technology. Firstly, during this period new
technologies are expensive, crude and imperfect
[@rosenberg1994exploring; @wilson201281]. A simple comparison between
Web 1.0 and Web 3.0 applications clearly illustrates this
[@tranos2020social]: while a static website facilitates one-way
information dissemination, a platform like `github` enables cooperation
between users and the creation of new information, meaning, and even
knowledge [@faraj2016special; @barassi2012does]. During this period the
performance of a new technology is the main attraction and not the cost
to access and use it [@wilson2011lessons]. There is a broader
theoretical discussion in the literature about the early adoption
motives. As summarised by @perkins2005international, on one hand,
epidemic models highlight the role of interpersonal contacts as a way
for new technologies to diffuse. On the other hand, economic models
underline the importance of heterogeneity. Different firms have
different structures and business plans, which define the potential
economic returns of the adoption of a new technology and, therefore, the
choice to adopt a new technology becomes an individual option. From a
broader and evolutionary perspective, initial conditions are essential
for the creation and evolution of path-dependent technological
development trajectories [@neffke2011regions; @simmie2014new]. This
argument is even more relevant when the focus is on digital technologies
because of the commonly found lag between investment and economic
returns as reflected in the Solow paradox [@acemoglu2014return;
@brynjolfsson2018artificial].

[^1]: <https://www.theguardian.com/global/2015/mar/22/web-browser-came-back-haunt-microsoft>

<!-- data -->
The data used here depict the active engagement with the Web as they
contain geolocated and time-stamped counts of commercial websites from
the UK. Instead of adopting metrics reflecting the passive engagement
with the Internet such as Internet adoption and speeds [e.g.
@blank2018local; @destefano2022fuzzy], this paper observes yearly counts
of geolocated, commercial websites, which reflect the creation and 
maintenance of such websites. It needs to be highlighted here that although 
there are various types of underpinning web technologies -- namely markup,
stylesheet, client- and server-side coding languages as well as
different web browsers and other related programming language -- the
focus here is on the diffusion of the Web as an overarching technology
instead of more specific web technologies. For a such a detailed and 
cross-country technological diffusion study, see @PAPAGIANNIDIS2015308.

<!-- description of the methods and results -->
To analyse these data, a diverse set of methods are employed. Firstly,
cumulative adoption curves are estimated for the UK as a whole and for
the Local Authority Districts (LAD), illustrating, for the first time,
such S-shaped diffusion patterns for local areas. The exploratory spatial data
analysis offers some first indications of the importance of the different 
spatial diffusion mechanisms in shaping the diffusion of the Web in the UK.
Importantly, the analysis adopts two distinct geographical scales: the LAD and 
the most detailed scale of UK census areas, the Output Areas (OA). Then, a
novel methodological framework based on Machine Learning (ML) algorithms
is developed to test the role of such mechanisms at these two scales. The 
results reveal, for the first time, the importance of different spatial 
diffusion mechanisms in the diffusion of a digital technology at very granular
geographical scales. The analysis also reveals that the diffusion of the Web 
is characterised by spatial heterogeneity, well-established spatial patterns 
and both stability and volatility.

<!-- contents -->
The structure of the paper goes as follows. @sec-litreview reviews the
mechanisms that shape the diffusion of a new technology and sets up the two 
hypotheses that are empirically tested in @sec-results. @sec-datamethods 
describes the data and methods used in this paper. The discussion of the 
results and the conclusions of the paper are offered in @sec-conclusions.

# The mechanisms that shape the diffusion of a new technology {#sec-litreview}

<!-- The fundamentals of technology diffusion -->
The diffusion of new technologies and innovations is driven by three
different mechanisms [@grubler1990rise; @morrill2020spatial]: (i)
distance and proximity, (ii) urban hierarchies in the form of a
centrifugal forces as new technologies diffuse from core to periphery,
and (iii) over time the cumulative level of adoption follows an S-shaped
pattern.
<!-- R2 addition -->
Although some studies focused on other, case-specific drivers -- for example 
@lengyel2020role on urban scaling and @bednarz2020pulled on regional demand --
the literature is in agreement that these are the three fundamental diffusion 
mechanisms. 
<!-- neighbourhood -->
@hagerstrand1968innovation was the first to identify diffusion as a
geographical process. The starting point was the idea that diffusion is
based on passing information through social networks, which themselves
tend to be geographically bounded. Hence, he identified a
'neighbourhood' mechanism of how information, and consequently,
innovations and new technologies diffuse. This perspective draws
similarities with epidemics: new technologies and innovations, just like
pathogens, spread because of contagion and, consequently, proximity and
exposure [@hivner2003facilitating].

<!-- hierarchy -->
@hagerstrand1968innovation also emphasised the role of the top-down
hierarchy of urban systems and highlighted that new technologies and
innovations are firstly adopted in larger cities and then diffuse to
second tier ones. This is a sequential instead of a simultaneous
process, which resembles the 'lead-lag' spatial acceleration effect in
market research [@bento2018time; @PERES201091]. Schmidt's Law
empirically illustrates a similar pattern [@grubler1990rise]. Core and
highly agglomerated regions is where new technologies are invented and
commercially deployed. This is where the first adopters tend to be
based. Then, technologies spread to the rim and eventually to the
periphery. Although adoption pace might be higher when new technologies
finally arrive to the periphery, the saturation levels may never
reach the ones in the core because of the lack of infrastructure or
other necessary institutions [@leibowicz2016representing].

<!-- time -->
Diffusion is naturally a dynamic process and Hägerstrand recognised the
role time plays in the diffusion of innovations and new technologies: an
early-pioneering period, a middle fast accelerating period and a final
saturation one [@morrill2020spatial]. This was further explored by
@rogers2010diffusion, who identified five types of adopters of new
technologies and innovations: innovators, early adopters, early
majority, late majority and laggards. Hence, plotting the cumulative
adoption of a new technology over time often leads to an S-shaped line
plot [@grubler1990rise].

Although the above processes have been studied in innovation studies,
the distancing of economic geographers from research questions about the
diffusion of new technologies and innovations resulted to a knowledge gap
regarding the explanatory power of these mechanisms at granular, local
scales [@ding2010modeling]. As discussed in @sec-introduction, such 
spatially explicit knowledge can directly inform local technology
and infrastructure policies. Indeed, while the spatial dimension is
present in the following studies, the geographical level of analysis is
often too coarse to ascertain the role of the above mechanisms at the
local scale. @beardsell1999spatial studied the evolution of the computer
industry in 317 US metropolitan areas during the 1977-1992 period using
employment data. Their analysis indicated that the relative size
distribution holds for urban computer employment and also urban
heterogeneity is essential in explaining this distribution. In a recent
study, @bednarz2020pulled focused on wind turbines and modelled their
spatial diffusion across 402 German regions during 1970-2015. Their key
finding is that local demand than local supply was the main driving
factor. @haller2011determinants employed firm panel data for Ireland
during 2003-2005 and illustrated that when firms are located in a region
or industry with high share of firms having a website, they are more
likely to have a website too. At a global scale @perkins2011internet
explored whether the adoption of previous communication technologies
that is mail, telegrams and telephones were shaped by similar
socioeconomic factors as the Internet. Their results indicated common
patterns regarding the drivers behind the adoption of different
communication technologies. @feldman2015rknowledge focused on inventions
and used patent data to map how the the rDNA technology diffused among
inventors across 366 US metropolitan areas. @ding2010modeling modelled
the spatial diffusion of mobile telecommunications across the highest
level of administrative division in China (29 regions in total). Their
analysis indicated that socioeconomic characteristics are important
determinants of the timing, speed and the level of mobile diffusion
within China. @bakher2013diffusion compared different growth curves
as a means to model the diffusion of online shopping in Australia during
the 1998-2009 period, but they did not consider any spatial mechanisms.
At a global scale @PAPAGIANNIDIS2015308 modelled the diffusion of
different web technologies and practices across countries. Using similar
but less extensive data as the one used here, their analysis illustrated
how the diffusion of different web technologies and practices follows an
S-shaped pattern as well as the different diffusion rates of the
different technologies and practices. The most detailed consideration of
the spatial footprint of the diffusion of a new technology comes from
@lengyel2020role, who analysed the adoption and churn of a Hungarian
online social network. Their results were in agreement with the early
theoretical and empirical contributions reviewed here: assortativity,
urban scaling and distance are the key drivers of spatial diffusion.
Based on the above, the main empirical hypothesis that is tested in this paper
states:

> H~1~: The three diffusion mechanisms namely distance, urban hierarchy
and the S-shaped pattern of the cumulative level of adoption shaped
the diffusion of the commercial Web in the UK even at granular geographical 
scales.

<!-- EARLY/LATE adoption-->
Understanding the diffusion of a new technology in multiple, detailed spatial
scales is of interest to economic geography as potential local economic 
benefits may occur because of concentration of early adopters or through 
technological leapfrogging. On the one hand, early adopters are often rewarded 
because of their attitudes towards new technologies and innovations. 
@rogers2010diffusion identifies them as 'knowledgeable risk takers' and 
@griliches1957 as 'profit maximisers' [@ding2010modeling]. On the other hand,
latecomers can adopt and benefit from new technologies that have been developed
elsewhere without incurring the hefty initial R&D costs [@teece2008firm]. At a 
global scale @perkins2005international explored whether the diffusion rate of 
new technologies is driven by a latecomer advantage and the engagement with the 
global economy via foreign direct investments and trade. Their results 
illustrate that indeed latecomers and developing countries experience diffusion 
of new technologies more rapidly than early adopters and developed countries. 
At a regional scale, previous research highlighted the long term and 
sustained productivity benefit of the early adopters of web technologies in 
the UK [@tranosuk]. Although the early adoption and leapfrogging literature has 
not paid much attention on cities and regions [@yu2018sustainability], the 
potential local economic benefits justify the need to investigate the
dynamics of the web diffusion in different scales in the UK. Hence, the second 
empirical hypothesis states:

> H~2~: The diffusion of the commercial Web is shaped by both early adoption 
and latecomer effects.

These two hypotheses are empirically tested in @sec-results employing 
the data and methods described in the next section.

<!-- Technological diffusion, which is by definition an aggregated process, -->
<!-- can be discussed in parallel with individual adoption mechanisms. On the -->
<!-- one hand,  -->
<!-- On the other hand, @perkins2011internet attribute diffusion -->
<!-- to two processes: (i) epidemic-like mechanisms, which are governed by -->
<!-- distance, proximity and social interactions, and (ii) economic -->
<!-- mechanisms as new innovations are adopted by users as they become more -->
<!-- profitable, valueable and useful. -->

# Data and Methods {#sec-datamethods}

<!-- main data -->
To capture the diffusion of the commercial Web, a website density metric is
developed for two different geographical scales that are discussed later in 
this section. The main data are counts of commercial websites that are 
calculated using data from the Internet Archive[^2] and, specifically, the 
JISC UK Web Domain Dataset [@ukwebarchive]. The Internet Archive is one of the 
most complete and oldest archive of webpages in the world operating since
1996 [@ainsworth2011much; @holzmann2016dawn]. It is a web crawler, which
discovers webpages by following the hyperlinks of every webpage it
archives. This dataset, which is curated by the British Library,
contains all the archived webpages from the UK country code top level
domain (ccTLD - .uk) from the 1996–2012 period. In essence, this is a
long list of 2.5 billion URLs of archived webpages including also the
archival timestamp.

[^2]: [https://archive.org/](https://archive.org/).

Instead of using the whole .uk ccTLD, this paper focuses on its
commercial subset, the .co.uk second level domain (SLD). This choice
is aligned with the topic of the paper -- the diffusion of the Web for 
commercial functions -- and decreases the heterogeneity of the web data as such 
commercial websites have specific aims: they are used to diffuse information, 
support online transactions and share opinions [@THELWALL2000441; 
@blazquez2018big]. Although a UK company can adopt a generic TLD such as .com 
and these cases escape the data used here, such omissions should not affect the
validity of the results given the popularity of the .uk ccTLD
[@tranosuk]: UK consumers prefer to visit a .uk website when they are
searching for products or services [@hope]; and anecdotal evidence
indicates that during the first half of 2000, three .co.uk domains were
registered every minute [@oecd_coms]. Importantly, previous studies
illustrated that .co.uk is the most popular UK SLD [@tranosuk]. Henceforward 
any reference to websites will refer to commercial -- .co.uk -- websites.

The text from these webpages was scanned using a regular expression
(regex) to identify strings of text which resemble UK postcodes and one
fifth of them included a mention to a postcode [@BL2013geo]. This
information allows the geolocation of the data and the creation of the
LAD and OA counts. To prevent false positives, the resulted postcodes
were matched against official postcode lookup files [@ons].

The data cleaning process includes an aggregation step, through which
the archived webpages are aggregated to the parent websites. This
website reconstruction process lend itself to the construction of
*website* instead of *webpage* density metrics. Websites tend to
represent specific organisations or entities, and, arguably, are more
meaningful observational units than webpages, which can ignore the
upstream dependency of the website they belong to. Based on the
following example, all three webpages are part of the same website
(<http://www.website.co.uk>). Only the overall website and not the
nested webpages are considered for the counts as otherwise the density
metrics would have been biased towards large, place specific, websites.
It is important to note here that websites commonly contain place references, 
with previous research estimating this share to be around 70% 
[@hill2009georeferencing].

-   <http://www.website.co.uk/webpage_a> B15 2TT

-   <http://www.website.co.uk/webpage_b> BS8 1TH

-   <http://www.website.co.uk/webpage_c> B15 2TT

\noindent What is challenging is that this aggregation approach, which
has been used elsewhere [@tranosuk; @shoreditch] may lead to websites
with multiple postcodes. As per the above example,
[www.website.co.uk](http://www.website.co.uk) includes two unique
postcodes: B15 2TT and BS8 1SS. The distribution of the number of
postcodes per website for 2000 is presented in Table \ref{f2000}, which
illustrates a wide range. At the left end of the distribution, there are
websites anchored to a unique location (72% of all the reconstructed
websites in 2000), which may represent a small company with a single
trading location. At the right end, there are websites with thousands of
different postcodes. Considering the time period of the analysis, such
cases can represent directories which used to be popular in the
pre-search engines early times of the commercial Internet as well as
real estate websites [@tranosuk]. Previous research which used the same 
methodology for a small subset of these data illustrates the robustness of the 
above approach: its results matched in terms of accuracy past rounds of 
research based on extensive qualitative work, interviews and observational 
studies for the same study area [@shoreditch].

The analysis is firstly conducted on websites, which only contain one
unique postcode. As a robustness check, the analysis is replicated for
an extended subset of websites, which include up to 10 unique postcodes
to capture commercial websites which point to multiple locations. They
are geolocated by equally attaching them to these locations. This
extended sample includes 94 percent of all the archived websites in
2000 (see Table \ref{f2000}).

```{r eval=T, echo=FALSE, results='asis'}

# load 2000 co.uk
# path.2000 <- "C:/Users/nw19521/OneDrive - University of Bristol/projects/archive/nuts/all2000couk.csv"
# path.2000 <- "/Users/nw19521/Library/CloudStorage/OneDrive-UniversityofBristol/projects/archive/nuts/all2000couk.csv"
path.2000 <- paste0(path, "/data/temp/all2000couk.csv")

all2000.duplicates <- data.table::fread(path.2000) #
# dim(all2000.duplicates) # 3336162
# it includes duplicated URLs: if one webpages includes multiple postcodes
# then it appears multiple times. This is ok for the nuts aggregation, but
# not for the frequencies
# This is only .co.uk

# one line for every host
all2000 <- unique(all2000.duplicates, by = c("host")) # 57897

# unique postcodes per website f table
f.websites.pc <- DescTools::Freq(all2000$V1, breaks = c(0, 1,2, 10,100,1000,10000,100000), ord = "desc")
f.websites.pc$level <- factor(f.websites.pc$level, levels = c("[0,1]","(1,2]", "(2,10]", "(10,100]",
                                                              "(100,1e+03]", "(1e+03,1e+04]",
                                                              "(1e+04,1e+05]"))
levels(f.websites.pc$level) <- c("(0,1]","(1,2]", "(2,10]", "(10,100]", "(100,1000]", "(1000,10000]", "(10000,100000]")

f.websites.pc <- f.websites.pc %>% 
  dplyr::select(-cumfreq) %>% 
  rename(Postcodes = level,
         F = freq,
         'F (%)' = perc,
         'cumulative F' = cumperc)

kable(f.websites.pc,
      format = "latex",
      digits = 3,
      booktabs = T,
      format.args = list(big.mark = ","),
      caption = "Number of unique postcodes per .co.uk website, 2000.\\label{f2000}") %>%
  kableExtra::footnote(general = "Tranos et al. 2021",
                       general_title = "Source: ",
                       footnote_as_chunk = T)  
   # kable_classic(full_width = F, html_font = "Cambria")
```

Another data cleaning step deals with some extreme outliers. Figure
\ref{correct} plots the website counts for the top 1000 postcodes. Some
obvious outliers can be observed for the 2002-2006 period for a handful
of postcodes, which can be attributed to link farms [@BL2013links]. The
website counts for these postcodes (SE24 9HP, CV8 2ED, GL16 7YA, CW1
6GL, M28 2SL, DE21 7BF), which in 2004 or 2005 had more than 1,000
websites pointing to them, were replaced with predicted values based on
a simple panel regression model with postcode fixed effects and yearly
dummy variables. These postcodes refer to a residential area with a
small park in South London, residential areas outside Warwick and
Manchester, a rural area in the south border of England and Wales and
small business parks in Crew and Derby. Hence, it is difficult to
believe that these postcodes genuinely hosted such a large amounts of
websites, which were later extensively declined. To put the magnitude of
the data imputation into perspective, this process affected 6 out of the
557,808 postcodes included in the data. As @sec-results illustrates,
this led to a small increase of the model predictive power.

<!-- Multi-scale -->
To empirically test the two hypotheses stated in @sec-litreview, the website 
density metric was calculated for two different scales the Local Authority
Districts (LAD) and the Output Areas (OA). The former is the lowest
administrative unit in the UK and there are 374 LAD. The latter is the smallest
census-based geographical unit and there are c. 230,000 of them in the UK. To 
create the website density metric, the yearly website counts at the LAD level 
are standardised by the number of firms in LAD to avoid biases associated with 
LAD hosting a large number of firms. Given that there is no such readily 
available statistic for the OA, the actual OA level counts are used. However, 
because of the the consistent spatial definition of OA (they host 40-250
households),[^3] website counts in OA are interpreted as a density metric too.

[^3]: <https://www.ons.gov.uk/methodology/geography/ukgeographies/statisticalgeographies>

<!-- S -->
These website density metrics are used for the different stages of the analysis. 
Starting with the first hypothesis (H~1~), to test whether the S-curve 
accurately describes the cumulative adoption of the Web at a local scale, the
following logistic function (Equation \ref{eq:s}) is estimated for the whole of 
the UK and then for each LAD separately:

\begin{align}
y = k /(1 + e^{-(t-t_{o})})\label{eq:s}
\end{align}

\noindent $k$ is the asymptote or, in other words, the saturation level,
$b$ the overall growth rate, and $t_{0}$ the *inflection point* of
maximum growth at $k/2$, where the logistic function is symmetrical
[@wilson201281]. The $t_0$ of each LAD is compared against the $t_0$ of
the UK to delineate whether a LAD reached that point faster or slower
than the country average. This heterogeneity is discussed further when testing 
the second hypothesis (H~2~). Importantly, an accuracy criterion was imposed
and only LAD with $R^2 > 0.9$ were included in this analysis. To
estimate Equation \ref{eq:s} a self-starting logistic growth model was
employed using the `nls` and `SSlogis` functions in `R`. This logistic
growth model depicts the initial period of slow diffusion of a new
technology, which is followed by the fast, exponential growth stage to
then eventually slow down and saturate [@wilson201281;
@grubler1999dynamics].

<!-- **minimum R2 = 95%** in @wilson201281, see also @grubler1990rise -->
<!-- The literature usually uses the saturation level as the asymptote. I am -->
<!-- using the total number of websites as we cannot compute a rate. -->

```{r echo=FALSE, message=FALSE, fig.cap="\\label{correct}Yearly website counts per postcode (top 1000) before (top) and after (down) data imputation", out.height="70%"}
path.image1 <- paste0(path, "/outputs/pc_year_1.png")
path.image2 <- paste0(path, "/outputs/pc_year_1_corrected.png")

p1 <- ggdraw() + draw_image(path.image1)  
p2 <- ggdraw() + draw_image(path.image2)

p1 + plot_spacer() + p2 + plot_layout(ncol = 2)#, widths = c(1, -0.8, 1))
```

<!-- ESDA -->
Next, exploratory spatial data analysis provides some first insights of whether 
the other two drivers of spatial diffusion -- namely neighbourhood and 
hierarchy -- underpin the diffusion of the Web in the UK. To capture the former 
and following @ding2010modeling the Moran's I and the Local Index of Spatial
Association (LISA) are estimated for the website density. To capture the
hierarchy mechanism, the Gini coefficient -- an established metric of
inequality -- is calculated. All the above are computed and plotted
longitudinally both for the LAD and the OA.

<!-- conceptual model -->
To formally test H~1~ a modelling framework is developed to assess the 
above diffusion mechanisms. The overarching aim is to build a model that can 
test how well these three mechanisms can predict website density:

\begin{align}
Website\,Density_{t} \sim Hierarchy_{t-1} + Neighbourhood_{t-1} + S\text-curve_{t}\label{eq:rf.generic}
\end{align}

<!-- <!-- conceptual model --> 
<!-- The models include variables that capture the three processes that the -->
<!-- relevant literature and the descriptive analysis highlighted. Namely, -->
<!-- the models capture: (i) a hierarchical mechanism with diffusion running -->
<!-- from main centres to secondary ones, (ii) a neighborhood mechanism -->
<!-- according to which diffusion first hits nearby locations, and (iii) the -->
<!-- rather canonical pattern of diffusion over time as reflected in the -->
<!-- S-shaped pattern in the cumulative level of adoption. -->

<!-- detailed features -->
To capture the hierarchical mechanism the model includes as predictors a
one year lag of website density in London, the largest city in the UK, and a
one year lag of the website density in the nearest city. Both variables clearly
illustrate the urban hierarchy in the UK. Especially for the OA models, a 
third predictor is added to capture hierarchy at a much more granular scale:
a one year lag of the website density in the the nearest retail centre [^4]. As
the literature illustrates, retail centres are key components of urban systems
[@dennis2002central; @jones2021reframing]. Due to the small sizes of the retail 
centres, the latter is only relevant for and included in the OA models and not 
in the LAD models. By including this predictor, the model is better tuned 
towards the spatial characteristic of the OA scale. Hence, it increases the 
robustness of the model.

In addition, the model includes the distance to London, the 
nearest city and the nearest retail centre. The underlying logic is that the 
level of website adoption in a spatial unit depends on the level of the 
adoption in places further up in the urban hierarchy the previous year. The
inclusion of the distance variables incorporates spatial structure into
the hierarchy argument. To depict the neighbourhood mechanism, the
website density of the neighbouring spatial units in the previous year
is employed. This is calculated using rook continuity. Again, the
underpinning rationale is that the level of web adoption within a
spatial unit depends on the level of web adoption in the neighbouring
spatial units the year before. This represents the 'hitting nearby
locations first' argument. Therefore, the spatial and temporal lag of
the website density in LAD and OA is calculated. Lastly, the temporal
pattern of the cumulative adoption is captured by a time trend variable.
Hence, the model will follow the following generic form (Eq.
\ref{eq:model}):

[^4]: The retail centres data have been obtained by the Consumer Data Research Centre, an ESRC data investment, under project ID CDRC 498-01, ES/L011840/1; ES/L011891/1: <https://data.cdrc.ac.uk/dataset/retail-centre-boundaries-and-open-indicators>.

\begin{align} \label{eq:model}
Website\,Density_{t} \sim Distance\,London +
Website\,density\,London_{t-1} +\notag\\
Distance\,Nearest\,City +
Website\,density\,Nearest\,City_{t-1} +\notag\\
Distance\,Nearest\,Retail_{i} +
Website\,density\,Nearest\,Retail_{t-1} +\notag\\
W*\, Website\,density_{t-1} +\notag\\ 
year_{t}
\end{align}

<!-- 4 models -->
Specifically, four different models are estimated. Firstly, all the data points 
for the OA and LAD are utilised in order to build two models and assess their
capacity to predict the adoption of the Web at the two different scales. These 
two models will reveal the predictive capacity of the diffusion mechanisms and 
also demonstrate how the importance of such variables changes across
different scales. The next two sets of models are again trained at these
two scales: OA and LAD. However, instead of using all the data points,
the OA and the LAD from one of the twelve UK regions are held out for
the model training. Then, the trained model is used to predict website
density for the OA or the LAD of the held-out region. This process takes
place recursively for all twelve UK regions. The regional differences in
the predictive capacity of the different samples will reveal how
dissimilar these spatial process are across regions and, importantly, at
different scales.

<!-- RF -->
Equation \ref{eq:model} is estimated using Random Forest
(RF). This is a popular ML algorithm for both regression and
classification problems [@biau2012analysis]. It was introduced by
@breiman2001random and has become a go-to data science tool. RF can
effectively handle skewed distributions and outliers, model non-linear
relationships, require minimal hyperparameter tuning, exhibit low
sensitivity to these parameters, and have relatively short training
times [@Caruana2008; @liaw2002classification; @yan2020using]. These
attributes match well with the website density data characteristics
including skewness especially for the OA. Also, the large data size (c.
230k data points for each of the 17 years) calls for fast training
times. Importantly, RF predictions tend to be more accurate than those
from single regression trees and outperform Ordinary Least Squares in
out-of-sample predictions, even with moderate-sized training data and a
small number of predictors [@mullainathan2017machine; @athey2019machine;
@sulaiman2011intelligent; @pourebrahim2019trip; @biau2012analysis].

<!-- RF -->
RF is a tree-based ensemble learning algorithm [@breiman2001random]. It
begins by generating random samples of the training data, which are then
used to grow regression trees to predict the dependent variable. Data
points and predictors are randomly sampled for the different trees. The
trees are trained in parallel using their own bootstrapped samples of
the training data. A crucial feature of RF is their ability not to
overfit, meaning they can generalize well to unseen test data. While
each tree may overfit individually, the ensemble of trees does not
because the errors of individual trees are averaged, reducing the
overall variance and preventing overfitting [@last2002improving]. For
regression problems, RF predictions are made by averaging the
predictions of all decision trees.

RF have been widely employed to address regression research problems.
@pourebrahim2019trip combined a spatial interaction modelling framework
with ML algorithms including RF to predict commuting flows in New York
City. @sinha2019assessing advocated for adopting spatial ensemble
learning approaches, such as RF, to model spatial data with high
autocorrelation and heterogeneity. @creditspatial predicted employment
density in Los Angeles using spatially explicit RF.
@guns2014recommending used RF to build a recommendation system for
research collaborations. @ren2019predicting trained RF to predict the
socio-economic status of cities using various online and mobility
predictors. @tranos2023using utilised hyperlinks data and RF to make
out-of-sample predictions of interregional trade. @zhou2023geography
employed such a framework to assess whether key predictors of obesity
differ across English cities.

<!-- CV -->
It needs to be highlighted here that the cross validation (CV) for all
models is spatially and temporally sensitive. Instead of using 10 random
and space- and time-agnostic folds, the `CAST` package is employed which
allows holding back data points from specific years and spatial units (regions)
and use them for testing in order to estimate the model performance
[@meyer2018improving].

>>>>

---
title: "A multi-scale story of the diffusion of a new technology: the Web"
date: last-modified
format:
  elsevier-pdf:
    keep-tex: true
    text: |
      \usepackage[demo]{graphicx} % "demo" option just for this example
      \usepackage{subcaption}
    journal:
      name: Journal of Economic Geography
      model: 3p
      cite-style: authoryear
editor: 
  markdown: 
    wrap: 72
---

@@

<!-- This file is for creating a stand alone appendix. It was created from tranos2025.qmd after deleting everything but the appendix part and adding an equation set counter.-->

<!-- To remove the corresponding author footnote, the appendix.tex should be manually tweaked: delete the \cortext[cor1]{Corresponding author} part. -->

\setcounter{section}{0}
\renewcommand{\thesection}{\Alph{section}}
\setcounter{table}{0}
\renewcommand{\thetable}{A\arabic{table}}
\setcounter{figure}{0}
\renewcommand{\thefigure}{A\arabic{figure}}
\setcounter{equation}{0}
\renewcommand{\theequation}{A\arabic{equation}}
